\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events

### 2023-09-22 paper/main.tex
spell check

We designed a naturalistic paradigm for exposing participants to scenarios
where the past and future were equally unobserved. We asked our participants to
watch a series of movie segments drawn from a character-driven dramatic
television show. Across the conditions and trials in the experiment,
participants made free-form text responses to either retrodict what had
happened in the previous segment, predict what would happen in the next
segment, or recall what happened in the just-watched segment. We used manual
annotations and sentence-level natural language processing models to
characterize participants' responses. To foreshadow our results, we found that
participants were overall better at retrodicting the past than predicting the
future. This appeared to be driven by two main factors. First, characters more
often referred to past events than future (e.g., planned) events, and this
influenced participants' responses. Second, associations and dependencies
between temporally adjacent events enabled participants to form estimates about
nearby events (e.g., to a just-watched scene or a past or future event
referenced in an observed conversation). We also ran a pre-registered
replication study to confirm that these findings generalized to another
television show and group of participants. Finally, we ran a meta analysis
using natural language processing to estimate the prevelance of references to
past and future events in hundreds of millions of dialogues drawn from
television shows, popular movies, novels, and written and spoken natural
conversations. Taken together, our work reveals a temporal asymmetry in how
observations of other humans’ behaviors inform us about the past versus the
future.

@@

Controlling for temporal distance (lag), past and future events that story
characters referenced in their conversations were associated with higher hit
rates than unreferenced events in our main experiment (uncued retrodiction: OR
= 12.70, $Z = 10.94$, $p < 0.001$, CI: 8.06 to 20.03; uncued prediction: OR =
8.29, $Z = 6.83$, $p < 0.001$, CI: 4.52 to 15.20; Fig.~\ref{fig:result3}E).
This indicates that participants' responses are at least partially influenced
by the characters' conversations. To estimate the contributions of characters’
references on hit rates, we computed the difference in hit rates between all
events (which comprised both referenced and unreferenced events) and
unreferenced events, as a function of lag. These differences exhibited a
temporal asymmetry in favor of retrodiction (Figs.~\ref{fig:result3}C). This
indicates that the asymmetries in participants' retrodictions versus
predictions are also at least partially influenced by the characters'
conversations. However, these temporal asymmetries in participants'
retrodictions and predictions persisted even for events that characters never
referenced in their conversations (hit rates of uncued retrodicted versus
predicted unreferenced events: OR = 2.00, $Z = 2.40$, $p = 0.02$, CI: 1.14 to
3.51; Fig.~\ref{fig:result3}D). When we further separated the unreferenced
events into onscreen events and offscreen events, we found that these
asymmetries held only for the onscreen events (onscreen: OR = 2.65, $Z = 2.59$,
$p = 0.01$, CI: 1.27 to 5.54; offscreen: OR = 1.50, $Z = 0.91$, $p = 0.36$, CI:
0.63 to 3.62). We found similar patterns in our replication experiment
(Fig.~\characterRefs; hit rates of uncued retrodictions for referenced events:
OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX; uncued predictions for
referenced events: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX; hit rates of
uncued retrodcitions for \textit{unreferenced} events: OR = XXX, $Z = XXX$, $p
= XXX$, CI: XXX to XXX; for predicted events: OR = XXX, $Z = XXX$, $p = XXX$,
CI: XXX to XXX). Taken together, these analyses suggest that asymmetries in the
number of references characters make to past and future events partially (but
not entirely) explain why participants tend to retrodict the past further and
more accurately than they predict the future.

@@

\paragraph{Replication experiment.} The stimuli used in our replication
experiment were segments of the first episodie of the Netflix television show
\textit{The Chair}, Season 1. \textbf{JRM NOTE: Describe the show, like you did
for Why Women Kill.} The mean length of the segments was XXX min (range
XXX--XXX min). As for the stimulus we used in our main experiment, we chose
this stimulus for our replication experiment for its linear storytelling
(again, within each storyline) and its realistic depections of everyday events.
\textit{JRM NOTE: The plots were focused on... (fill in something analogous to
what you wrote for the main experiment stimulus...)}

>>>>

We designed a naturalistic paradigm for exposing participants to scenarios
where the past and future were equally unobserved. We asked our participants to
watch a series of movie segments drawn from a character-driven dramatic
television show. Across the conditions and trials in the experiment,
participants made free-form text responses to either retrodict what had
happened in the previous segment, predict what would happen in the next
segment, or recall what happened in the just-watched segment. We used manual
annotations and sentence-level natural language processing models to
characterize participants' responses. To foreshadow our results, we found that
participants were overall better at retrodicting the past than predicting the
future. This appeared to be driven by two main factors. First, characters more
often referred to past events than future (e.g., planned) events, and this
influenced participants' responses. Second, associations and dependencies
between temporally adjacent events enabled participants to form estimates about
nearby events (e.g., to a just-watched scene or a past or future event
referenced in an observed conversation). We also ran a pre-registered
replication study to confirm that these findings generalized to another
television show and group of participants. Finally, we ran a meta analysis
using natural language processing to estimate the prevalence of references to
past and future events in hundreds of millions of dialogues drawn from
television shows, popular movies, novels, and written and spoken natural
conversations. Taken together, our work reveals a temporal asymmetry in how
observations of other humans’ behaviors inform us about the past versus the
future.

@@

Controlling for temporal distance (lag), past and future events that story
characters referenced in their conversations were associated with higher hit
rates than unreferenced events in our main experiment (uncued retrodiction: OR
= 12.70, $Z = 10.94$, $p < 0.001$, CI: 8.06 to 20.03; uncued prediction: OR =
8.29, $Z = 6.83$, $p < 0.001$, CI: 4.52 to 15.20; Fig.~\ref{fig:result3}E).
This indicates that participants' responses are at least partially influenced
by the characters' conversations. To estimate the contributions of characters’
references on hit rates, we computed the difference in hit rates between all
events (which comprised both referenced and unreferenced events) and
unreferenced events, as a function of lag. These differences exhibited a
temporal asymmetry in favor of retrodiction (Figs.~\ref{fig:result3}C). This
indicates that the asymmetries in participants' retrodictions versus
predictions are also at least partially influenced by the characters'
conversations. However, these temporal asymmetries in participants'
retrodictions and predictions persisted even for events that characters never
referenced in their conversations (hit rates of uncued retrodicted versus
predicted unreferenced events: OR = 2.00, $Z = 2.40$, $p = 0.02$, CI: 1.14 to
3.51; Fig.~\ref{fig:result3}D). When we further separated the unreferenced
events into onscreen events and offscreen events, we found that these
asymmetries held only for the onscreen events (onscreen: OR = 2.65, $Z = 2.59$,
$p = 0.01$, CI: 1.27 to 5.54; offscreen: OR = 1.50, $Z = 0.91$, $p = 0.36$, CI:
0.63 to 3.62). We found similar patterns in our replication experiment
(Fig.~\characterRefs; hit rates of uncued retrodictions for referenced events:
OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX; uncued predictions for
referenced events: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX; hit rates of
uncued retrodictions for \textit{unreferenced} events: OR = XXX, $Z = XXX$, $p
= XXX$, CI: XXX to XXX; for predicted events: OR = XXX, $Z = XXX$, $p = XXX$,
CI: XXX to XXX). Taken together, these analyses suggest that asymmetries in the
number of references characters make to past and future events partially (but
not entirely) explain why participants tend to retrodict the past further and
more accurately than they predict the future.

@@

\paragraph{Replication experiment.} The stimuli used in our replication
experiment were segments of the first episode of the Netflix television show
\textit{The Chair}, Season 1. \textbf{JRM NOTE: Describe the show, like you did
for Why Women Kill.} The mean length of the segments was XXX min (range
XXX--XXX min). As for the stimulus we used in our main experiment, we chose
this stimulus for our replication experiment for its linear storytelling
(again, within each storyline) and its realistic depictions of everyday events.
\textit{JRM NOTE: The plots were focused on... (fill in something analogous to
what you wrote for the main experiment stimulus...)}
